Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2018-05-08 articles/formalism/sections/kernel_discussion.tex
updated kernel figures and the updated text to indicate the correct mapping between kernels and their Q/U and E/B interpretation

\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_rker_rad_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_iker_rad_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_rker_con_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_iker_con_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_rker_D_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_iker_D_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_rker_I_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_iker_I_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure[$ \textrm{Re} \left(\mathcal{M}_{G} \right) $]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_rker_rad_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Img} \left(\mathcal{M}_{G} \right) $]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_iker_rad_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Re}  \left(\mathcal{M}_{B} \right) $]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_rker_con_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Img} \left(\mathcal{M}_{B} \right) $]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_iker_con_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Re} \left(\mathcal{D} \right)$]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_rker_D_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Img} \left(\mathcal{D} \right)$]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_iker_D_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Re} \left(\mathcal{I} \right)$]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_rker_I_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Img} \left(\mathcal{I} \right)$]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_iker_I_lat0_lon90.pdf}}\hspace{-2mm}
\caption{This panel of figure depicts the real and imaginary parts of the real space kernels $\mathcal{M}_{G}$, $\mathcal{M}_{B}$, $\mathcal{D}$ \& $\mathcal{I}$ respectively. These kernels have been evaluated with the band limit: $\ell \in [2,192]$ and sampled at the Healpix resolution parameter NSIDE=2048 for visual appeal. The size of each panel is approximately $16^{\circ} \times 16^{\circ}$ and the grid lines are marked at 2 degree separations. The black circles denotes the position of the center around which the kernels have been evaluated while the black star marks the location of the north galactic pole. The four rows depict the kernels at different location on the sphere and the galactic coordinates of the central pixel are specified in each panel.}
%from top to bottom rows are as follows $[b,\ell] = [0^{\circ},0^{\circ}], [87^{\circ},0^{\circ}], [87^{\circ},30^{\circ}], [80^{\circ},30^{\circ}], [0^{\circ},90^{\circ}]$.}
\label{fig:vis_kernel}
 \end{figure}
%

@@

We compute the Euler angles $(\alpha, \beta, \gamma)$ given the angular coordinates of any two Healpix pixels and use these to evaluate the convolution and the radiating kernels. To provide an intuition for how these kernels vary as a function of position of the central pixel we depict in \fig{fig:vis_kernel} the respective kernels at a few different locations on the sphere.
While the kernels are evaluated in the band limit $\ell \in [2,192]$, for illustration these functions are sampled at a very high Healpix resolution parameter of NSIDE=2048. All the plots have been rotated such that the central location marked by the black circle are in the centre of the figure. The horizontal and vertical lines that pass through the central black circle mark the local latitude and longitude respectively.

The real and imaginary part of the kernel $\mathcal{M}_G$ are identical irrespective of changes in the galactic latitude and longitude of the central pixel. This is consistent with the Green's function interpretation of these kernels. Note that these functions are not distorted when a part of the domain overlaps with the poles, as can be seen in the first three rows of \fig{fig:vis_kernel}. Both these facts can be associated with the fact that this function does not depend on the Euler angle $\gamma$. From \eq{eq:qu2eb_convolution_explicit} and \eq{eq:eb2qu_convolution_explicit} it is clear that $\mathcal{M}_r$ and $\mathcal{M}_i$ can be interpreted in the following ways,
%\revisit{Recall that the angle $\gamma$ defines the rotation about the final z-axis which co-alligns the planar axes. Since the function $\mathcal{M}$ does not depend of this final orientation, it implies a rotational invariance of the fields constructed by convolving with this kernel.}
%
\begin{subequations}
\beqry
\bmat E= -\mathcal{M}_r \\ B =+\mathcal{M}_i  \emat  \leftarrow \bmat Q=\delta(\hat{n} - \hat{n}_j)\\ U=0 \emat ~~;~~ \bmat E= -\mathcal{M}_i \\ B = - \mathcal{M}_r  \emat  \leftarrow \bmat Q=0 \\ U=\delta(\hat{n} - \hat{n}_j) \emat \,,  \\
\bmat Q= -\mathcal{M}_r \\ U = -\mathcal{M}_i  \emat  \leftarrow \bmat E=\delta(\hat{n} - \hat{n}_j)\\ B=0 \emat ~~;~~ \bmat Q= +\mathcal{M}_i \\ U = - \mathcal{M}_r  \emat  \leftarrow \bmat E=0 \\ B=\delta(\hat{n} - \hat{n}_j) \emat \,.
\eeqry
\end{subequations}
%
The kernels $\mathcal{D}$ \& $\mathcal{I}$ vary significantly as a function of galactic latitude of the central pixel as seen in the last four columns of \fig{fig:vis_kernel}. These kernels show a two fold symmetry in the vicinity of the poles and this arises due to Euler angle $\gamma \approx 0$ here and therefore $e^{i2(\alpha \pm \gamma)} \approx e^{i2\alpha}$. Note that in this region, the azimuthal profile of the real and imaginary part of these kernels  is similar to $\mathcal{M}_r$ and $\mathcal{M}_i$ respectively.  This explains why the imaginary part of the band limited delta function $\mathcal{I}$ contributes just as much as the real part in these regions. On transiting to lower latitudes, $\mathcal{D}$ quickly transitions to having a four fold symmetry while $\mathcal{I}$ transitions to being dominated by the real part and behaves more like the conventional delta function. This transition can be most easily understood in the flat sky limit where $\gamma = -\alpha$ which leads to the resultant 4 fold symmetry seen for $\mathcal{D}$ owing to $e^{i2(\alpha - \gamma)} =e^{i4\alpha}$ and $\mathcal{I}$ being dominated by the real part owing to $e^{-i2(\alpha + \gamma)} =1 + i0$. Since the flat sky approximation has most validity in the proximity of the equator these limiting tendencies of the respective kernels are seen in the bottom row of \fig{fig:vis_kernel} which depict the kernels evaluated at the equator $b=0^{\circ}$. The middle two row depict the kernels evaluated at a latitudes of $b=87^{\circ}~\&~ 80^{\circ}$ and serve to indicate the rate of this transition. These kernels are invariant under changes in longitude of the central pixel, the latitude being held fixed, as one may have expected.

>>>>

\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_rker_rad_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_iker_rad_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_rker_con_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_iker_con_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_rker_D_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_iker_D_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_rker_I_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_iker_I_lat80_lon30.pdf}}\hspace{-2mm}
\subfigure[$ \textrm{Re} \left(\mathcal{M}_{G} \right) $]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_rker_rad_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Im} \left(\mathcal{M}_{G} \right) $]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_iker_rad_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Re}  \left(\mathcal{M}_{B}^* \right) $]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_rker_con_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Im} \left(\mathcal{M}_{B}^* \right) $]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2eb_iker_con_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Re} \left(\mathcal{D} \right)$]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_rker_D_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Im} \left(\mathcal{D} \right)$]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_iker_D_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Re} \left(\mathcal{I} \right)$]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_rker_I_lat0_lon90.pdf}}\hspace{-2mm}
\subfigure[$\textrm{Im} \left(\mathcal{I} \right)$]{\includegraphics[width=0.125\columnwidth]{new_kernel/qu2ebqu_iker_I_lat0_lon90.pdf}}\hspace{-2mm}
\caption{This panel of figure depicts the real and imaginary parts of the real space kernels $\mathcal{M}_{G}$, $\mathcal{M}_{B}$, $\mathcal{D}$ \& $\mathcal{I}$ respectively. These kernels have been evaluated with the band limit: $\ell \in [2,192]$ and sampled at the Healpix resolution parameter NSIDE=2048 for visual appeal. The size of each panel is approximately $16^{\circ} \times 16^{\circ}$ and the grid lines are marked at 2 degree separations. The black circles denotes the position of the center around which the kernels have been evaluated while the black star marks the location of the north galactic pole. The four rows depict the kernels at different location on the sphere and the galactic coordinates of the central pixel are specified in each panel.}
%from top to bottom rows are as follows $[b,\ell] = [0^{\circ},0^{\circ}], [87^{\circ},0^{\circ}], [87^{\circ},30^{\circ}], [80^{\circ},30^{\circ}], [0^{\circ},90^{\circ}]$.}
\label{fig:vis_kernel}
 \end{figure}
%

@@

We compute the Euler angles $(\alpha, \beta, \gamma)$ given the angular coordinates of any two Healpix pixels and use these to evaluate the convolution and radiation kernels. To provide an intuition for how these kernels vary as a function of position of the central pixel we depict in \fig{fig:vis_kernel} the respective kernels at a few different locations on the sphere.
While the kernels are evaluated in the band limit $\ell \in [2,192]$, for illustration these functions are sampled at a very high Healpix resolution parameter of NSIDE=2048. All the plots have been rotated such that the central location marked by the black circle are in the centre of the figure. The horizontal and vertical lines that pass through the central black circle mark the local latitude and longitude respectively.

The real and imaginary part of the kernel $\mathcal{M}_G$ are identical irrespective of changes in the galactic latitude and longitude of the central pixel. This is consistent with the Green's function interpretation of these kernels. Note that these functions are not distorted when a part of the domain overlaps with the poles, as can be seen in the first three rows of \fig{fig:vis_kernel}. Both these facts can be associated with the fact that this function does not depend on the Euler angle $\gamma$. From \eq{eq:qu2eb_convolution_explicit} and \eq{eq:eb2qu_convolution_explicit} it is clear that $\mathcal{M}_r$ and $\mathcal{M}_i$ can be interpreted in the following ways,
%\revisit{Recall that the angle $\gamma$ defines the rotation about the final z-axis which co-alligns the planar axes. Since the function $\mathcal{M}$ does not depend of this final orientation, it implies a rotational invariance of the fields constructed by convolving with this kernel.}
%
\begin{subequations}
\beqry
\bmat E= \textrm{Re}(\mathcal{M}_G) \\ B =\textrm{Im}(\mathcal{M}_G)  \emat  \leftarrow \bmat Q=\delta(\hat{n} - \hat{n}_q)\\ U=0 \emat ~;~ \bmat E= -\textrm{Im}(\mathcal{M}_G) \\ B = + \textrm{Re}(\mathcal{M}_G)  \emat  \leftarrow \bmat Q=0 \\ U=\delta(\hat{n} - \hat{n}_q) \emat \,,\\
\bmat Q= \textrm{Re}(\mathcal{M}^*_B) \\ U = \textrm{Im}(\mathcal{M}^*_B)  \emat  \leftarrow \bmat E=\delta(\hat{n} - \hat{n}_e)\\ B=0 \emat ~;~ \bmat Q= -\textrm{Im}(\mathcal{M}^*_B) \\ U = \textrm{Re}(\mathcal{M}^*_B)  \emat  \leftarrow \bmat E=0 \\ B=\delta(\hat{n} - \hat{n}_e) \emat \,.
\eeqry
\end{subequations}
%
The kernels $\mathcal{D}$ \& $\mathcal{I}$ vary significantly as a function of galactic latitude of the central pixel as seen in the last four columns of \fig{fig:vis_kernel}. These kernels show a two fold symmetry in the vicinity of the poles and this arises due to Euler angle $\gamma \approx 0$ here and therefore $e^{i2(\alpha \pm \gamma)} \approx e^{i2\alpha}$. Note that in this region, the azimuthal profile of the real and imaginary part of these kernels  is similar to $\mathcal{M}_r$ and $\mathcal{M}_i$ respectively.  This explains why the imaginary part of the band limited delta function $\mathcal{I}$ contributes just as much as the real part in these regions. On transiting to lower latitudes, $\mathcal{D}$ quickly transitions to having a four fold symmetry while $\mathcal{I}$ transitions to being dominated by the real part and behaves more like the conventional delta function. This transition can be most easily understood in the flat sky limit where $\gamma = -\alpha$ which leads to the resultant 4 fold symmetry seen for $\mathcal{D}$ owing to $e^{i2(\alpha - \gamma)} =e^{i4\alpha}$ and $\mathcal{I}$ being dominated by the real part owing to $e^{-i2(\alpha + \gamma)} =1 + i0$. Since the flat sky approximation has most validity in the proximity of the equator these limiting tendencies of the respective kernels are seen in the bottom row of \fig{fig:vis_kernel} which depict the kernels evaluated at the equator $b=0^{\circ}$. The middle two row depict the kernels evaluated at a latitudes of $b=87^{\circ}~\&~ 80^{\circ}$ and serve to indicate the rate of this transition. These kernels are invariant under changes in longitude of the central pixel, the latitude being held fixed, as one may have expected.

### 2018-05-08 articles/formalism/sections/real_space_formalism.tex
updated kernel figures and the updated text to indicate the correct mapping between kernels and their Q/U and E/B interpretation

&=& 0.5 \Bigg\lbrace \mathcal{I}^* \circ {}_{+2}X\pm \mathcal{D} \circ {}_{+2}X^* \Bigg\rbrace \,, \label{eq:qu2equbqu_convolution}

@@

where we have suppressed the explicit multipole dependence of functions ${}_{\mdi}f$ for brevity. \revisit{Comment on the radiating and convolution form of the operators.}

\comment{Recheck the math described below}
The operator $\mathcal{I}$ is a band limited version of the delta function ($\lim_{\ell \rightarrow \infty} \mathcal{I} = \delta(\hat{n}_i - \hat{n}_j)$) for spin-2 fields. When interpreted as a matrix it is a band limited version of the identity matrix. Though it has non-vanishing off diagonal elements ($\mathcal{I} \neq 0 $ when $\hat{n}_i \neq \hat{n}_j$) owing to the band limit, for all practical purposes $\mathcal{I}$ acts like an identity operator as is confirmed by the following set of identities: (i) $\mathcal{I}*\mathcal{I}=\mathcal{I}$ ; (ii) $\mathcal{D}*\mathcal{I}=\mathcal{D}$. Also $\mathcal{D}^*$ is the inverse of $\mathcal{D}$ in this band limited sense: $\mathcal{D}^**\mathcal{D}=\mathcal{I}$. It is useful to note that the operator $\mathcal{D}$ is a complex but symmetric matrix and $\mathcal{I}$ is an Hermitian operator. Using these properties of the operators $\mathcal{I}$ and $\mathcal{D}$, one can verify that the real space operators satisfy the following identities,
%
\begin{subequations}
\beqry
\bar O_E * \bar O_E &=& \bar O_E ~~;~~ \bar O_B * \bar O_B = \bar O_B \,, \\
\bar O_E*\bar O_B &=& 0 \,,\\
\bar O_E + \bar O_B &=& \mathcal{I} \,,
\eeqry
\end{subequations}
%
which are the real space analogues of \eq{eq:har_op_prop}. While testing the above stated identities one encounters terms like $\mathcal{D}*\mathcal{I}^*,\mathcal{I}^**\mathcal{I} \textrm{ and }  \mathcal{I}*\mathcal{I}^*$ which cannot be simply interpreted, but they always occur in pairs with opposite signs, hence exactly cancel each other.

Note that unlike in the harmonic case, the sum of the operators is not exactly an identity matrix. 
%\comment{Isn't it possible to do some real space operations which doesn't care for the band limit but still returns the required decomposition into the respective scalar modes ?} 
This non-exactness is representative of the loss of information resulting from making this transformation on the measured data with some imposed band limit. Forcing the sum of the operators to be exactly an identity matrix compromises the orthogonality property of the $\bar{O}_E$ \& $\bar{O}_B$ operators, which is exact.

>>>>

&=& 0.5 \Bigg\lbrace  {}_{+2}X \cdot \mathcal{I}^*_G \pm {}_{+2}X^* \cdot \mathcal{D}_G \Bigg\rbrace \hspace{.5cm }\textrm{\emph{   Radiation kernel}} \,, \\
&=& 0.5 \Bigg\lbrace \mathcal{I}_B \circ {}_{+2}X\pm \mathcal{D}_B \circ {}_{+2}X^* \Bigg\rbrace   \hspace{.5cm } \textrm{\emph {  Convolution kernel}} \,, \label{eq:qu2equbqu_convolution}

@@

where all the symbols have their usual meaning and the explicit multipole dependence of the real space operators has been suppressed for brevity. Note that when the operators $\mathcal{I}^*$ \& $\mathcal{D}$ are expressed in terms of the Euler angles $(\alpha_{qe},\beta_{qe},\gamma_{qe})$ they can be interpreted as the Greens functions and  we denote them as $\mathcal{I}^*_G$ \& $\mathcal{D}_G$. When expressed as function of Euler angles $(\alpha_{eq},\beta_{eq},\gamma_{eq})$ corresponding to the inverse rotations they can be interpreted as a some convolving beam and we denote them as $\mathcal{I}_B$ \& $\mathcal{D}_B$. Note that unlike in the case of the operators $\mm_G$ \& $\mm_B$ which have different shapes owing to their dependence on Euler angles $\alpha$ and $\gamma$ respectively, the operators $D_G$ \& $D_B$ are identical since $(\alpha_{qe}-\gamma_{qe}) = (\alpha_{eq}-\gamma_{eq})$, while $\mathcal{I}^*_{G}$ \& $\mathcal{I}_B$ are related by conjugation since  $(\alpha_{qe}+\gamma_{qe}) = -(\alpha_{eq}+\gamma_{eq})$.

The operator $\mathcal{I}$ is Hermitian and is a band limited version of the delta function owing to the identity $\lim_{\ell \rightarrow \infty} \mathcal{I} = \delta(\hat{n}_i - \hat{n}_j)$. For all practical purposes $\mathcal{I}$ acts like an identity operator as is confirmed by the following set of identities: (i) $\mathcal{I}*\mathcal{I}=\mathcal{I}$ ; (ii) $\mathcal{D}*\mathcal{I}=\mathcal{D}$. $\mathcal{D}$ is a complex but symmetric matrix and $\mathcal{D}^*$ is its inverse in this band limited sense: $\mathcal{D}^**\mathcal{D}=\mathcal{I}$. Using these properties of the operators $\mathcal{I}$ and $\mathcal{D}$, one can verify that the real space operators satisfy the following identities,
%
\begin{subequations}
\beqry
\bar O_E * \bar O_E &=& \bar O_E ~~;~~ \bar O_B * \bar O_B = \bar O_B \,, \\
\bar O_E*\bar O_B &=& 0 \,,\\
\bar O_E + \bar O_B &=& \mathcal{I} \,,
\eeqry
\end{subequations}
%
which are the real space analogues of \eq{eq:har_op_prop}. \footnote{While testing the above stated identities one encounters terms like $\mathcal{D}*\mathcal{I}^*,\mathcal{I}^**\mathcal{I} \textrm{ and }  \mathcal{I}*\mathcal{I}^*$ which cannot be simply interpreted but they always occur in pairs with opposite signs that exactly cancel each other.}

Note that unlike in the harmonic case, the sum of the operators is the band limited identity operator $\mathcal{I}$. This non-exactness is representative of the loss of information resulting from making this transformation on the measured data with some imposed band limit. Forcing the sum of the operators to be exactly an identity matrix compromises the orthogonality property of the $\bar{O}_E$ \& $\bar{O}_B$ operators which is exact and a more crucial property of the operators.
